Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-23 paper/paper.tex
Merge adequacy lemmas into a single one

Before we can prove the fundamental theorem and derive the
normalization property as its corollary, we need to additionally
formulate and prove the adequacy properties.

Let us first start with some auxiliary lemmas and definitions.
\begin{lemma}[Ext Wn]
  \label{lemma:extwn}
  If $[[wn (a var i)]]$, then $[[wn a]]$.
\end{lemma}
\begin{proof}
  By induction over the length of the reduction sequence in $[[wn (a
  var i)]]$. The conclusion follows from Lemma~\ref{lemma:parantirenaming} and
  \ref{lemma:parnenf}.
\end{proof}

@@

To state the adequacy lemmas, it is useful to first define the notion
of reducibility candidates (shortened as $CR$), predicates over sets
of lambda terms that capture the properties we need from our logical
relation.
\begin{definition}[CR]
  Let $[[S]]$ be a set of lambda terms. We say that
  \begin{itemize}
  \item $[[S]] \in CR_1 \iff $  $[[forall a, (#  wne a implies a in S #)]]$
  \item $[[S]] \in CR_2 \iff$ $[[forall a, (# a in S implies wn a #)]]$
  \item $[[S]] \in CR \iff $ $[[S]] \in CR_1$ and $[[S]] \in CR_2$
  \end{itemize}
\end{definition}

\begin{lemma}[Adequacy: $CR_1$ for type]
  \label{lemma:cr1ty}
  If $[[wne A]]$, then we have $[[Interp I i A {a | wne a }]]$.
\end{lemma}
\begin{proof}
  Immediately from \rref{I-Ne} and \rref{I-Par}.
\end{proof}

\begin{lemma}[Adequacy: $CR$ for interpreted sets]
  \label{lemma:crel}
  If $[[I]](j) \in CR$ for all $[[j < i]]$ and $[[Interp I i A S]]$, then $[[S]] \in CR$.
\end{lemma}
\begin{proof}
  By induction over the derivation of $[[Interp I i A S]]$. The base
  cases are all immediate.

  In the function case ($[[Pi A B]]$), we use the induction
  hypothesis to show that variables, which are special cases of
  neutral terms, semantically inhabit the input type $[[A]]$. We apply the
  function to an arbitrary variable and we can then use
  Lemma~\ref{lemma:extwn} to conclude the $CR_2$ property.

  To show
  the $CR_1$ property, we need to prove that given a neutral term $[[b]]$
  such that $[[wne b]]$,
  and a term $[[a]]$ that semantically inhabits $[[A]]$, $[[b a]]$ gives
  us a term that semantically inhabits $[[B {a}]]$. From the induction
  hypothesis, $[[A]]$ also satisfies $CR_2$ and therefore $[[wn a]]$
  holds. By Lemma~\ref{lemma:wnewn}, we have $[[wne b a]]$. By $CR_2$
  from the induction hypothesis, since every term that evaluates to
  some neutral form semantically inhabits $[[B {a}]]$. Done.
\end{proof}

\begin{lemma}[Adequacy: $CR_2$ for type]
  \label{lemma:cr2ty}
  If $[[I]](j) \in CR$ for all $[[j < i]]$ and $[[Interp I i A S]]$, then $[[wn A]]$ holds.
\end{lemma}
\begin{proof}
  By induction over the derivation of $[[Interp I i A S]]$. The
  function case uses Lemma~\ref{lemma:crel} and Lemma~\ref{lemma:extwn}.
\end{proof}

\begin{lemma}[Adequacy: $CR_2$ for types (Rec)]
  \label{lemma:cr2tyrec}
  If $[[InterpR i A S]]$, then $[[wn A]]$.
\end{lemma}
\begin{proof}
  By strong induction over $[[i]]$ and Lemma~\ref{lemma:cr2ty}.
\end{proof}

\begin{lemma}[Adequacy: $CR$ for terms (Rec)]
  \label{lemma:crelrec}
  If $[[InterpR i A S]]$, then $[[S]] \in CR$.
\end{lemma}
\begin{proof}
  After unfolding the definition of $[[InterpR i A S]]$, trivial by
  Lemmas~\ref{lemma:cr1ty}, \ref{lemma:crel} and \ref{lemma:cr2tyrec}.
\end{proof}

The formulation of the valuation and semantic well-typedness from
Figure~\ref{fig:semtyping} and the fundamental lemma remains
unchanged.
The proof of the fundamental lemma is still carried out by induction
over the typing derivation, where the additional neutral term related
cases are handled by Lemmas~\ref{lemma:cr2tyrec} and
\ref{lemma:crelrec}.

@@

Dependent types make the normalization proof slightly more complicated
as we also need to know that every type has a normal form.
Whereas the
$CR_1$ property for types (Lemma~\ref{lemma:cr1ty}) is directly
derivable, the $CR_2$ property for types (Lemma~\ref{lemma:cr2ty})
requires the $CR_1$ property about the interpreted sets
(Lemma~\ref{lemma:crel}) to be derived first.

>>>>

Before we can prove the fundamental theorem and derive the
normalization property as its corollary, we need to additionally
formulate and prove the adequacy property.

Let us first start with some auxiliary lemmas and definitions.
\begin{lemma}[Wn extensionality]
  \label{lemma:extwn}
  If $[[wn (a var i)]]$, then $[[wn a]]$.
\end{lemma}
\begin{proof}
  By induction over the length of the reduction sequence in $[[wn (a
  var i)]]$. The conclusion follows from Lemma~\ref{lemma:parantirenaming} and
  \ref{lemma:parnenf}.
\end{proof}

@@

To state the adequacy lemma, we first define the notion
of reducibility candidates (shortened as $CR$), predicates over sets
of lambda terms that capture the properties we need from our logical
relation.
\begin{definition}[CR]
  Let $[[S]]$ be a set of lambda terms. We say that
  \begin{itemize}
  \item $[[S]] \in CR_1 \iff $  $[[forall a, (#  wne a implies a in S #)]]$
  \item $[[S]] \in CR_2 \iff$ $[[forall a, (# a in S implies wn a #)]]$
  \item $[[S]] \in CR \iff $ $[[S]] \in CR_1$ and $[[S]] \in CR_2$
  \end{itemize}
\end{definition}

\yl{Merged all the separate adequacy related properties into a single
  property proved through nested induction. (The mechanization has
  been adjusted accordingly)}
We now state and prove the adequacy lemma. We give some more details
since we prove it through nested induction.
\begin{lemma}[Adequacy\footnote{\dotv{soundnessopen.v}{soundness}}]
  \label{lemma:adequacy}
  If $[[InterpR i A S]]$, then we have $[[S]] \in CR$.
\end{lemma}
\begin{proof}
  We start by strong induction over $[[i]]$. We are given the
  induction hypothesis that for all $[[j < i]]$, $[[InterpR j A S]]$
  implies $[[S]] \in CR$. Our goal is to show $[[InterpR i A S]]$
  implies $[[S]] \in CR$.

  By its definition in Figure~\ref{fig:logrelrec}, we have the
  equality $[[InterpR i A S]] = [[Interp I i A S]]$  where $[[I i]] :=
  [[{A | exists S , InterpR i A S}]]$.
  We then proceed by structural over the derivation of $[[Interp I i A
  S]]$. The only interesting cases are \rref{I-Pi} and \rref{I-Set}.
  The function case requires Lemmas~\ref{lemma:extwn} and
  \ref{lemma:wnewn} we have shown earlier.

  The \rref{I-Set} case is the most involved. We need to show that
  for all $[[j < i]]$, the set $[[{ A | exists S, InterpR j A S  }]]
  \in CR$. We immediately know that $[[{A | exists S, InterpR j A S
  }]] \in CR_1$ by \rref{I-Ne}. It remains to show that $[[{A | exists S, InterpR j A S
  }]] \in CR_2$, or equivalently, for all $[[A]]$, $[[InterpR j A S]]$
  implies $[[wn A]]$. Suppose $[[InterpR j A S]]$ for an arbitrary
  $[[A]]$. We have $[[InterpR j A S]]=[[Interp I j A S]]$ where $[[I]]$
  has the same definition from earlier but its domain restricted to
  numbers less than $[[j]]$. We perform another induction on
  $[[j]]$. All cases are trivial except for the case for
  \rref{I-Pi}. Our induction hypothesis immediately gives us $[[wn
  A]]$. To derive $[[wn Pi A B]]$, it remains to show $[[wn B]]$. We
  use the outermost induction hypothesis to show that $[[var 0]]$
  semantically inhabits $[[A]]$, from which we derive $[[wn B {var 0}]]$
  and conclude $[[wn B]]$ through antirenaming (Lemma~\ref{lemma:parantirenaming}).
\end{proof}

% \begin{lemma}[Adequacy: $CR_1$ for type]
%   \label{lemma:cr1ty}
%   If $[[wne A]]$, then we have $[[Interp I i A {a | wne a }]]$.
% \end{lemma}
% \begin{proof}
%   Immediately from \rref{I-Ne} and \rref{I-Par}.
% \end{proof}

% \begin{lemma}[Adequacy: $CR$ for interpreted sets]
%   \label{lemma:crel}
%   If $[[I]](j) \in CR$ for all $[[j < i]]$ and $[[Interp I i A S]]$, then $[[S]] \in CR$.
% \end{lemma}
% \begin{proof}
%   By induction over the derivation of $[[Interp I i A S]]$. The base
%   cases are all immediate.

%   In the function case ($[[Pi A B]]$), we use the induction
%   hypothesis to show that variables, which are special cases of
%   neutral terms, semantically inhabit the input type $[[A]]$. We apply the
%   function to an arbitrary variable and we can then use
%   Lemma~\ref{lemma:extwn} to conclude the $CR_2$ property.

%   To show
%   the $CR_1$ property, we need to prove that given a neutral term $[[b]]$
%   such that $[[wne b]]$,
%   and a term $[[a]]$ that semantically inhabits $[[A]]$, $[[b a]]$ gives
%   us a term that semantically inhabits $[[B {a}]]$. From the induction
%   hypothesis, $[[A]]$ also satisfies $CR_2$ and therefore $[[wn a]]$
%   holds. By Lemma~\ref{lemma:wnewn}, we have $[[wne b a]]$. By $CR_2$
%   from the induction hypothesis, since every term that evaluates to
%   some neutral form semantically inhabits $[[B {a}]]$. Done.
% \end{proof}

% \begin{lemma}[Adequacy: $CR_2$ for type]
%   \label{lemma:cr2ty}
%   If $[[I]](j) \in CR$ for all $[[j < i]]$ and $[[Interp I i A S]]$, then $[[wn A]]$ holds.
% \end{lemma}
% \begin{proof}
%   By induction over the derivation of $[[Interp I i A S]]$. The
%   function case uses Lemma~\ref{lemma:crel} and Lemma~\ref{lemma:extwn}.
% \end{proof}

% \begin{lemma}[Adequacy: $CR_2$ for types (Rec)]
%   \label{lemma:cr2tyrec}
%   If $[[InterpR i A S]]$, then $[[wn A]]$.
% \end{lemma}
% \begin{proof}
%   By strong induction over $[[i]]$ and Lemma~\ref{lemma:cr2ty}.
% \end{proof}

% \begin{lemma}[Adequacy: $CR$ for terms (Rec)]
%   \label{lemma:crelrec}
%   If $[[InterpR i A S]]$, then $[[S]] \in CR$.
% \end{lemma}
% \begin{proof}
%   After unfolding the definition of $[[InterpR i A S]]$, trivial by
%   Lemmas~\ref{lemma:cr1ty}, \ref{lemma:crel} and \ref{lemma:cr2tyrec}.
% \end{proof}

The formulation of the valuation and semantic well-typedness from
Figure~\ref{fig:semtyping} and the fundamental lemma remains
unchanged.
The proof of the fundamental lemma is still carried out by induction
over the typing derivation, where the additional neutral term related
cases are handled by Lemma~\ref{lemma:adequacy}, the adequacy property.

@@

Dependent types make the adequacy proof slightly more complicated
as we also need to know that every type has a normal form, not just
the terms.
This complicates our proof specifically in the \rref{I-Set} case for
Lemma~\ref{lemma:adequacy}.
